Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2024-11-21 docs/paper.qmd
include new cross correlations by period table

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities, eventually spurring the transformation of commodity markets from a collection of idiosyncratic markets into a more cohesive collection akin to an asset class. In this context, interconnectedness emerges as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicates higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-combined.   
We break down these average correlations by years in @tbl-correlations-years. For the full sample of assets, the average pairwise correlations remain in the 0.06–0.08 range from 1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point for the onset of financialisation, as noted in several studies [@baker_financialization_2021; @tang_index_2012]. We further observe that the US metals appear to have played a major role in this as the increase seems driven by the sector which average pairwise correlation more than doubles (0.49 vs 0.22) over the period.

@@

The average pairwise correlation for the entire set of sample assets increases further to 0.28 (@tbl-correlations-periods) indicating an intensification of the higher interdependence within the full cross-section of commodities prompted by financialisation. The result is consistent with the greater interdependence between equity and commodity markets over this period documented in @silvennoinen_financialisation_2013. The figure for the US traded section of the metals sector only progresses modestly to 0.59 after peaking at 0.63 in 2006 (@tbl-correlations-years). In contrast, the GB traded assets see a sharp increase to over 0.65  starting in 2007 up to a peak of 0.69 in 2010, and so does the agricultural sector with the sharpest increase in 2008 when the average pairwise correlation for the sector jumps to 0.30. The energy sector also shows a substantial progression starting in 2007 and peaking at 0.63 in 2008 before levelling back down over the rest of the period. The onset of the financial crisis seems therefore to have had an across the board impact on the interdependence of commodities in contrast with financialisation, much more closely related to the metals sector.

@@

The average pairwise correlations fall across the board, below the levels observed during financialisation, with the figure for the entire cross-section (@tbl-correlations-periods) dropping down to 0.12 from 0.28 over the crisis. This substantial decline, that does not seem to have been widely noted, appears to begin in the latter part of the crisis period, around 2012 (@tbl-correlations-years), before levelling down over the whole post-crisis period.

@@

```{r}
#| label: tbl-correlations-periods
#| tbl-cap: "This table shows the average return pairwise correlation coefficients calculated independently across the whole cross-section of commodity assets considered in the study as well as across countries and commodity sectors and sub-sectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). For each period, the results are calculated independently over the entire period as well as over phases of aggregate backwardation and contango. Aggregate CHP construction and corresponding regime definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results and the taxonomy for the commodity assets considered is detailed in \\@sec-taxonomy."

@@

```{r}
#| label: tbl-correlations-years
#| tbl-cap: "This table shows the average return pairwise correlation coefficients calculated independently across the whole cross-section of commodity assets considered in the study as well as across countries and commodity sectors and sub-sectors by year for the whole period of interest (1997-2018). The results are discussed in \\@sec-results while \\@sec-taxonomy details the taxonomy for the commodity assets considered."

>>>>

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities, eventually spurring the transformation of commodity markets from a collection of idiosyncratic markets into a more cohesive collection akin to an asset class. In this context, interconnectedness emerges as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-inner-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicates higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-combined.   
We break down these average correlations by years in @tbl-correlations-inner-years. For the full sample of assets, the average pairwise correlations remain in the 0.06–0.08 range from 1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point for the onset of financialisation, as noted in several studies [@baker_financialization_2021; @tang_index_2012]. We further observe that the US metals appear to have played a major role in this as the increase seems driven by the sector which average pairwise correlation more than doubles (0.49 vs 0.22) over the period.

@@

The average pairwise correlation for the entire set of sample assets increases further to 0.28 (@tbl-correlations-inner-periods) indicating an intensification of the higher interdependence within the full cross-section of commodities prompted by financialisation. The result is consistent with the greater interdependence between equity and commodity markets over this period documented in @silvennoinen_financialisation_2013. The figure for the US traded section of the metals sector only progresses modestly to 0.59 after peaking at 0.63 in 2006 (@tbl-correlations-inner-years). In contrast, the GB traded assets see a sharp increase to over 0.65  starting in 2007 up to a peak of 0.69 in 2010, and so does the agricultural sector with the sharpest increase in 2008 when the average pairwise correlation for the sector jumps to 0.30. The energy sector also shows a substantial progression starting in 2007 and peaking at 0.63 in 2008 before levelling back down over the rest of the period. The onset of the financial crisis seems therefore to have had an across the board impact on the interdependence of commodities in contrast with financialisation, much more closely related to the metals sector.

@@

The average pairwise correlations fall across the board, below the levels observed during financialisation, with the figure for the entire cross-section (@tbl-correlations-inner-periods) dropping down to 0.12 from 0.28 over the crisis. This substantial decline, that does not seem to have been widely noted, appears to begin in the latter part of the crisis period, around 2012 (@tbl-correlations-inner-years), before levelling down over the whole post-crisis period.

@@

```{r}
#| label: tbl-correlations-inner-periods
#| tbl-cap: "This table shows the average return pairwise correlation coefficients calculated independently across the whole cross-section of commodity assets considered in the study as well as within countries and commodity sectors and sub-sectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). For each period, the results are calculated independently over the entire period as well as over phases of aggregate backwardation and contango. Aggregate CHP construction and corresponding regime definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results and the taxonomy for the commodity assets considered is detailed in \\@sec-taxonomy."

@@

```{r}
#| label: tbl-correlations-cross-periods
#| tbl-cap: "This table shows the return pairwise correlation coefficients between two equally weighted country portfolios (US, GB) with the first formed from the twenty four US commodities and the second from the six GB traded assets. The table further shows the average return pairwise correlation coefficients between equally weighted US sector portfolios (agriculturals, energy, metals) as well as between equally weighted US sub-sector portfolios (agriculturals: gains, livestock, softs; energy: gas, petroleum; metals: base, precious). The results are independently calculated for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). For each period, the results are further calculated independently over the entire period as well as over phases of aggregate backwardation and contango. Aggregate CHP construction and corresponding regime definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results and the taxonomy for the commodity assets considered is detailed in \\@sec-taxonomy."

colnames <- c("portfolios", "regime", periods)
align <- make_series_of_repeated_chars(list(l = 1L, r = 5L))

make_table("correlations - cross - periods", c("portfolios", "regime"), colnames, align, 1L:2L, 3L:6L)
```

```{r}
#| label: tbl-correlations-inner-years
#| tbl-cap: "While \\@tbl-correlations-inner-periods shows results calculated independently for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018), this stable shows the results of the same analysis calculated by year over the entire sample period (1997-2018). The results are discussed in \\@sec-results while \\@sec-taxonomy details the taxonomy for the commodity assets considered."

### 2024-11-21 docs/paper.tex
include new cross correlations by period table

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities, eventually spurring the transformation of commodity markets
from a collection of idiosyncratic markets into a more cohesive
collection akin to an asset class. In this context, interconnectedness
emerges as a crucial aspect that warrants thorough examination.\\
We strive to address this issue by first examining the average returns
pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicates higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We break down these average correlations by years in
Table~\ref{tbl-correlations-years}. For the full sample of assets, the
average pairwise correlations remain in the 0.06--0.08 range from 1998
through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This
suggests 2004 as a distinct break point for the onset of
financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

The average pairwise correlation for the entire set of sample assets
increases further to 0.28 (Table~\ref{tbl-correlations-periods})
indicating an intensification of the higher interdependence within the
full cross-section of commodities prompted by financialisation. The
result is consistent with the greater interdependence between equity and
commodity markets over this period documented in
\citet{silvennoinen_financialisation_2013}. The figure for the US traded
section of the metals sector only progresses modestly to 0.59 after
peaking at 0.63 in 2006 (Table~\ref{tbl-correlations-years}). In
contrast, the GB traded assets see a sharp increase to over 0.65
starting in 2007 up to a peak of 0.69 in 2010, and so does the
agricultural sector with the sharpest increase in 2008 when the average
pairwise correlation for the sector jumps to 0.30. The energy sector
also shows a substantial progression starting in 2007 and peaking at
0.63 in 2008 before levelling back down over the rest of the period. The
onset of the financial crisis seems therefore to have had an across the
board impact on the interdependence of commodities in contrast with
financialisation, much more closely related to the metals sector.

@@

The average pairwise correlations fall across the board, below the
levels observed during financialisation, with the figure for the entire
cross-section (Table~\ref{tbl-correlations-periods}) dropping down to
0.12 from 0.28 over the crisis. This substantial decline, that does not
seem to have been widely noted, appears to begin in the latter part of
the crisis period, around 2012 (Table~\ref{tbl-correlations-years}),
before levelling down over the whole post-crisis period.

@@

\caption{\label{tbl-correlations-periods}This table shows the average
return pairwise correlation coefficients calculated independently across
the whole cross-section of commodity assets considered in the study as
well as across countries and commodity sectors and sub-sectors for the
four periods of interest (past: 1997-2003; financialisation: 2004-2008;
crisis: 2008-2013; post-crisis: 2013-2018). For each period, the results
are calculated independently over the entire period as well as over
phases of aggregate backwardation and contango. Aggregate CHP
construction and corresponding regime definitions are discussed in
Section~\ref{sec-methods} while the results are discussed in
Section~\ref{sec-results} and the taxonomy for the commodity assets
considered is detailed in Section~\ref{sec-taxonomy}.}

@@

@@

\caption{\label{tbl-correlations-years}This table shows the average
return pairwise correlation coefficients calculated independently across
the whole cross-section of commodity assets considered in the study as
well as across countries and commodity sectors and sub-sectors by year
for the whole period of interest (1997-2018). The results are discussed
in Section~\ref{sec-results} while Section~\ref{sec-taxonomy} details
the taxonomy for the commodity assets considered.}

>>>>

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities, eventually spurring the transformation of commodity markets
from a collection of idiosyncratic markets into a more cohesive
collection akin to an asset class. In this context, interconnectedness
emerges as a crucial aspect that warrants thorough examination.\\
We strive to address this issue by first examining the average returns
pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector.
Table~\ref{tbl-correlations-inner-periods} shows the results for the
four periods of interest with the results reported for the entire period
as well as for periods of \textbf{backwardation} and \textbf{contango}.
For the entire sample the average pairwise correlation jumps from 0.07
in the first period to 0.16 over financialisation. This finding
indicates higher interdependence within the full cross-section of
commodities starting with the financialisation period, consistent with
the empirical findings in \citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We break down these average correlations by years in
Table~\ref{tbl-correlations-inner-years}. For the full sample of assets,
the average pairwise correlations remain in the 0.06--0.08 range from
1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004.
This suggests 2004 as a distinct break point for the onset of
financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

The average pairwise correlation for the entire set of sample assets
increases further to 0.28 (Table~\ref{tbl-correlations-inner-periods})
indicating an intensification of the higher interdependence within the
full cross-section of commodities prompted by financialisation. The
result is consistent with the greater interdependence between equity and
commodity markets over this period documented in
\citet{silvennoinen_financialisation_2013}. The figure for the US traded
section of the metals sector only progresses modestly to 0.59 after
peaking at 0.63 in 2006 (Table~\ref{tbl-correlations-inner-years}). In
contrast, the GB traded assets see a sharp increase to over 0.65
starting in 2007 up to a peak of 0.69 in 2010, and so does the
agricultural sector with the sharpest increase in 2008 when the average
pairwise correlation for the sector jumps to 0.30. The energy sector
also shows a substantial progression starting in 2007 and peaking at
0.63 in 2008 before levelling back down over the rest of the period. The
onset of the financial crisis seems therefore to have had an across the
board impact on the interdependence of commodities in contrast with
financialisation, much more closely related to the metals sector.

@@

The average pairwise correlations fall across the board, below the
levels observed during financialisation, with the figure for the entire
cross-section (Table~\ref{tbl-correlations-inner-periods}) dropping down
to 0.12 from 0.28 over the crisis. This substantial decline, that does
not seem to have been widely noted, appears to begin in the latter part
of the crisis period, around 2012
(Table~\ref{tbl-correlations-inner-years}), before levelling down over
the whole post-crisis period.

@@

\caption{\label{tbl-correlations-inner-periods}This table shows the
average return pairwise correlation coefficients calculated
independently across the whole cross-section of commodity assets
considered in the study as well as within countries and commodity
sectors and sub-sectors for the four periods of interest (past:
1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis:
2013-2018). For each period, the results are calculated independently
over the entire period as well as over phases of aggregate backwardation
and contango. Aggregate CHP construction and corresponding regime
definitions are discussed in Section~\ref{sec-methods} while the results
are discussed in Section~\ref{sec-results} and the taxonomy for the
commodity assets considered is detailed in Section~\ref{sec-taxonomy}.}

@@

\begin{longtable}[t]{>{}l>{}r>{}r>{}r>{}r>{}r}

\caption{\label{tbl-correlations-cross-periods}This table shows the
return pairwise correlation coefficients between two equally weighted
country portfolios (US, GB) with the first formed from the twenty four
US commodities and the second from the six GB traded assets. The table
further shows the average return pairwise correlation coefficients
between equally weighted US sector portfolios (agriculturals, energy,
metals) as well as between equally weighted US sub-sector portfolios
(agriculturals: gains, livestock, softs; energy: gas, petroleum; metals:
base, precious). The results are independently calculated for the four
periods of interest (past: 1997-2003; financialisation: 2004-2008;
crisis: 2008-2013; post-crisis: 2013-2018). For each period, the results
are further calculated independently over the entire period as well as
over phases of aggregate backwardation and contango. Aggregate CHP
construction and corresponding regime definitions are discussed in
Section~\ref{sec-methods} while the results are discussed in
Section~\ref{sec-results} and the taxonomy for the commodity assets
considered is detailed in Section~\ref{sec-taxonomy}.}

\toprule
\textcolor{black}{\textbf{portfolios}} & \textcolor{black}{\textbf{regime}} & \textcolor{black}{\textbf{past}} & \textcolor{black}{\textbf{financialisation}} & \textcolor{black}{\textbf{crisis}} & \textcolor{black}{\textbf{post-crisis}}\\
\midrule
\textbf{\cellcolor{gray!10}{countries}} & \textbf{\cellcolor{gray!10}{whole period}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{-0.0053}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{-0.0024}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0024}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{-0.0028}}\\
\textbf{} & \textbf{backwardation} & \textcolor[HTML]{4285f4}{0.0157} & \textcolor[HTML]{4285f4}{-0.0041} & \textcolor[HTML]{4285f4}{0.0106} & \textcolor[HTML]{4285f4}{-0.0127}\\
\textbf{\cellcolor{gray!10}{}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{-0.0100}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0078}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{-0.0183}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{-0.0050}}\\
\textbf{sectors} & \textbf{whole period} & \textcolor[HTML]{4285f4}{0.0496} & \textcolor[HTML]{4285f4}{0.0073} & \textcolor[HTML]{4285f4}{0.0637} & \textcolor[HTML]{4285f4}{0.0116}\\
\textbf{\cellcolor{gray!10}{}} & \textbf{\cellcolor{gray!10}{backwardation}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.1527}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0183}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.2099}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0902}}\\
\addlinespace
\textbf{} & \textbf{contango} & \textcolor[HTML]{4285f4}{0.0223} & \textcolor[HTML]{4285f4}{0.0240} & \textcolor[HTML]{4285f4}{-0.0141} & \textcolor[HTML]{4285f4}{0.0024}\\
\textbf{\cellcolor{gray!10}{subsectors}} & \textbf{\cellcolor{gray!10}{whole period}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0421}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0631}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0799}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0778}}\\
\textbf{} & \textbf{backwardation} & \textcolor[HTML]{4285f4}{0.1166} & \textcolor[HTML]{4285f4}{0.0201} & \textcolor[HTML]{4285f4}{0.0633} & \textcolor[HTML]{4285f4}{0.0727}\\
\textbf{\cellcolor{gray!10}{}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0297}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.1164}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0889}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{0.0812}}\\
\bottomrule

@@

\caption{\label{tbl-correlations-inner-years}While
Table~\ref{tbl-correlations-inner-periods} shows results calculated
independently for the four periods of interest (past: 1997-2003;
financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018),
this stable shows the results of the same analysis calculated by year
over the entire sample period (1997-2018). The results are discussed in
Section~\ref{sec-results} while Section~\ref{sec-taxonomy} details the
taxonomy for the commodity assets considered.}
